---
title: "Chinese homonym meaning-level AoA analysis"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Libraries
```{r libraries, message=FALSE, warning=FALSE}
# Load necessary libraries
library(ggplot2)
library(dplyr)
library(tidyr)
library(readr)
library(readxl)
library(purrr)
library(stringr)
library(tidyverse)
library(arrow)
library(rstatix)
library(emmeans)
library(ggpubr)
library(ez)
library(xtable)
library(lme4)
library(lmerTest)
library(ggstatsplot)
library(ggsci)
library(patchwork)

```

## Experiment 1: Data screening and preprocessing (rating RT)
```{r data screening and preprocessing, message=FALSE, warning=FALSE}
# set working directory to project folder (update as needed)
setwd("~/Documents/meaning-specific-AoA-of-562-Chinese-homonyms/data")

# Load data
df_all <- read_csv("Exp1rating_HPM_data.csv")  

# remove NA rows in Stimulus_RESP and Explist, and convert Stimulus_RESP and Stimulus_RT to numeric
df_all <- df_all %>%
  filter(!is.na(Explist)) %>%
  filter(!is.na(Stimulus_RESP)) %>%
  mutate(RESP = as.numeric(Stimulus_RESP)) %>%
  mutate(RT = as.numeric(Stimulus_RT))
```

### Data screening: remove extremely fast responses (RT < 500 ms)
```{r inspect_raw_data}
# report the number and percentage of RT values that is less than 500 ms for each subject
Less500RT = df_all %>%
  group_by(Subject) %>%
  summarize(count_fast_RT = sum(RT < 500, na.rm = TRUE),
             total_RT = n(),
             pct_fast_RT = (count_fast_RT / total_RT) * 100,
             .groups = "drop") %>%
  arrange(desc(pct_fast_RT)) %>%
  print()

# define Subject is subj
subj <- 35

gghistogram(df_all %>% filter(Subject== subj), x="RESP", bins=7)

# report the mean RT for each response for subj
df_all %>%
  filter(Subject == subj) %>%
  group_by(RESP) %>%
  summarize(mean_RT = mean(RT, na.rm = TRUE), .groups = "drop") %>%
  print()

# remove subject 35 for further analysis
df_all <- df_all %>% filter(Subject != 35)
```
Subject 35 appears to press button 4 immediately. We remove this subject from the analysis.

### Data screening: remove RT outliers

We compute subject-specific median and median absolute deviation, then flag the reaction time as outliers if
`abs(logRT - sbj_logRT_median) > 2.5 * sbj_mad(logRT)`, following the guidelines in Leys et al. (2013).

> Leys, C., Ley, C., Klein, O., Bernard, P., & Licata, L. (2013). Detecting outliers: Do not use standard deviation around the mean, use absolute deviation around the median. Journal of Experimental Social Psychology, 49(4), 764–766. https://doi.org/10.1016/j.jesp.2013.03.013

```{r remove_RT_outliers}
# For each participant, a trial was flagged as an outlier if |logRT - median(logRT)| > 2.5 × MAD(logRT)
df_all <- df_all %>%
  group_by(Subject) %>%
  mutate(
    logRT = log(RT),
    sbj_logRT_median = median(logRT, na.rm = TRUE),
    sbj_mad = mad(logRT, na.rm = TRUE),
    RT_outlier = abs(logRT - sbj_logRT_median) > 2.5 * sbj_mad
  ) %>%
  ungroup()

# reprot the percentage of rows removed as outliers for each subject
outlier_percentage <- df_all %>%
  group_by(Subject) %>%
  summarize(
    total_rows = n(),
    outliers_removed = sum(RT_outlier, na.rm = TRUE),
    outlier_pct = (outliers_removed / total_rows) * 100,
    .groups = "drop"
  ) %>%
  filter(outliers_removed > 0) %>%
  arrange(desc(outlier_pct))
print(outlier_percentage)  


# report the number and percentage of outliers removed
total_rows <- nrow(df_all)
outliers_removed <- sum(df_all$RT_outlier)
outlier_pct <- (outliers_removed / total_rows) * 100
message(glue::glue("Total rows: {total_rows}, Outliers removed: {outliers_removed}, Percentage of outliers removed: {round(outlier_pct, 2)}%"))

# report the number and percentage of outliers for each word type
outlier_type <-df_all %>%
  group_by(word_type) %>%
  summarise(
    n = n(),
    out_n = sum(RT_outlier == 1, na.rm = TRUE),
    out_pct = 100*out_n/n
  )
print(outlier_type)

# Remove outliers
df_HPM <- df_all %>%
  filter(!RT_outlier)

# reprot the number of rows after outlier removal
print(glue::glue("Rows after removing RT outliers: {nrow(df_HPM)}"))
```

### statistics on outliers by word type
Since word type may affect the likelihood of outliers, we fit a mixed-effects logistic regression model to examine the effect of word type on the probability of a trial being an outlier, with random intercepts for subjects and items.
```{r outlier statistics}
# convert word_type, item_id, Subject to factor if not already
if (!is.factor(df_all$word_type)) {
  df_all <- df_all %>%
    mutate(word_type = factor(word_type, levels = c("1","2","3"), labels = c("homonym", "polyseme", "monoseme")))
}
if (!is.factor(df_all$item_id)) {
  df_all <- df_all %>%
    mutate(item_id = factor(item_id))
}
if (!is.factor(df_all$Subject)) {
  df_all <- df_all %>%
    mutate(Subject = factor(Subject))
}
# convert RT, RESP, Explist to numeric if it is not already
if (!is.numeric(df_all$RT)) {
  df_all <- df_all %>%
    mutate(RT = as.numeric(as.character(RT)))
}
if (!is.numeric(df_all$RESP)) {
  df_all <- df_all %>%
    mutate(RESP = as.numeric(as.character(RESP)))
}
if (!is.numeric(df_all$Explist)) {
  df_all <- df_all %>%
    mutate(Explist = as.numeric(as.character(Explist)))
}

m_out_ri <- glmer(
  RT_outlier ~ word_type + (1 | Subject) + (1 | item_id),
  data = df_all,
  family = binomial,
  control = glmerControl(optimizer = "bobyqa")
)
summary(m_out_ri)
emmeans(m_out_ri, ~ word_type, type="response")
```

## Experiment 1: AoA ratings by word type
```{r data preparation for modeling}
# convert word_type, item_id, Subject to factor if not already
if (!is.factor(df_HPM$word_type)) {
  df_HPM <- df_HPM %>%
    mutate(word_type = factor(word_type, levels = c("1","2","3"), labels = c("homonym", "polyseme", "monoseme")))
}
if (!is.factor(df_HPM$item_id)) {
  df_HPM <- df_HPM %>%
    mutate(item_id = factor(item_id))
}
if (!is.factor(df_HPM$Subject)) {
  df_HPM <- df_HPM %>%
    mutate(Subject = factor(Subject))
}
# convert RT, RESP, Explist to numeric if it is not already
if (!is.numeric(df_HPM$RT)) {
  df_HPM <- df_HPM %>%
    mutate(RT = as.numeric(as.character(RT)))
}
if (!is.numeric(df_HPM$RESP)) {
  df_HPM <- df_HPM %>%
    mutate(RESP = as.numeric(as.character(RESP)))
}
if (!is.numeric(df_HPM$Explist)) {
  df_HPM <- df_HPM %>%
    mutate(Explist = as.numeric(as.character(Explist)))
}
```


```{r linear mixed model on RESP, message=FALSE, warning=FALSE}
# Main LMM for with outlier removal
m1 <- lmer(RESP ~ word_type + (1 + word_type | Subject) + (1 | item_id),
           data = df_HPM,
           control = lmerControl(optimizer = "bobyqa"))
summary(m1)
emmeans(m1, pairwise ~ word_type, adjust = "holm")

# Main LMM without outlier removal
m2 <- lmer(RESP ~ word_type + (1 + word_type | Subject) + (1 | item_id),
           data = df_all,
           control = lmerControl(optimizer = "bobyqa"))
summary(m2)
emmeans(m2, pairwise ~ word_type, adjust = "holm")
```

```{r}
# plot histogram of RESP by word_type
ggplot(df_HPM, aes(x = RESP, fill = word_type)) +
  geom_histogram(binwidth = 1, width = 0.7,
                 position = position_dodge(width = 0.9)) +
  labs(title = "Distribution of AoA ratings by word type", x = "Response", y = "Count", fill = "Word type") +
  theme_minimal()
```


## Experiment 1: Rating latency as a function of AoA rating and word type
Do higher AoA ratings (larger RESP) take longer to produce for a subject, and does that RESP→RT relationship differ by word_type?

```{r linear mixed model on RT, message=FALSE, warning=FALSE}
# Within-subject effect: on trials where a participant gives a higher-than-their-usual AoA rating, are they slower?
# Between-subject effect: participants who generally give higher AoA ratings, are they slower overall?
d2 <- df_HPM %>%
  mutate(
    logRT = log(RT),
    trial_z = as.numeric(scale(Explist)),
    # within- and between-subject decomposition of RESP
    RESP_mean_subj = ave(RESP, Subject, FUN = mean),
    RESP_w = RESP - RESP_mean_subj,
    RESP_b = as.numeric(scale(RESP_mean_subj))  # optional between-subject term
  )

m_rt <- lmer(
  logRT ~ RESP_w * word_type + trial_z + RESP_b +
    (1 + RESP_w | Subject) +
    (1 | item_id),
  data = d2,
  control = lmerControl(optimizer = "bobyqa")
)

summary(m_rt)

# Simple slopes: how much logRT changes per 1-point increase in RESP (within-subject),
# separately by word type
emtrends(m_rt, ~ word_type, var = "RESP_w")
pairs(emtrends(m_rt, ~ word_type, var = "RESP_w"), adjust = "holm")
# extract 95% CIs for the simple slopes
confint(m_rt, method="Wald")
```


```{r plot RT by RESP and word_type}
# compute means and then draw bars with labels and adjusted y-axis
plot_data <- d2 %>%
  # ensure word_type is labelled for plotting
  mutate(word_type = factor(word_type,
                            levels = c("1","2","3","homonym","polyseme","monoseme"),
                            labels = c("homonym","polyseme","monoseme","homonym","polyseme","monoseme"))) %>%
  group_by(RESP, word_type) %>%
  summarize(mean_logRT = mean(logRT, na.rm = TRUE), .groups = "drop")

ggplot(plot_data, aes(x = factor(RESP), y = mean_logRT, fill = word_type)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  geom_text(aes(label = round(mean_logRT, 2)),
            position = position_dodge(width = 0.8),
            vjust = -0.5, size = 3) +
  labs(title = "Mean log-transformed RT by AoA rating and word type",
       x = "RESP", y = "Average logRT", fill = "word_type") +
  coord_cartesian(ylim = c(6, 8)) +
  scale_y_continuous(breaks = seq(6, 8, 0.5)) +
  theme_minimal()

```


## Experiment 2: Anchor reproducibility and data quality
```{r load data and cleaning, warning=FALSE}
# set working directory to project folder (update as needed)
setwd("~/Documents/meaning-specific-AoA-of-562-Chinese-homonyms/data")

# read
df_anchors <- read_csv("anchor_trials_in_both_experiments.csv",
              col_types = cols(
                subject = col_integer(),
                item_id = col_integer(),
                headword = col_character(),
                Response = col_double(),
                RT_ms = col_double(),
                anchor = col_character(),
                list = col_character()

              ))

# basic cleaning
df_anchors <- df_anchors %>%
  mutate(
    anchor = factor(anchor, levels = c("exp1", "exp2")),
    list = factor(list, levels = c("List1", "List2")),
    item_id = factor(item_id),
    subject = factor(subject),
    Response = as.numeric(Response),
    RT_ms = as.numeric(RT_ms)
  )
```


```{r Item-level Exp1 vs Exp2 comparison}
item_sum <- df_anchors %>%
  group_by(anchor, item_id, headword) %>%
  summarise(
    n = n(),
    mean = mean(Response, na.rm = TRUE),
    sd = sd(Response, na.rm = TRUE),
    median = median(Response, na.rm = TRUE),
    iqr = IQR(Response, na.rm = TRUE),
    .groups = "drop"
  )

item_comp <- item_sum %>%
  select(anchor, item_id, headword, n, mean, median, sd, iqr) %>%
  pivot_wider(names_from = anchor, values_from = c(n, mean, median, sd, iqr)) %>%
  mutate(
    diff_mean = mean_exp2 - mean_exp1,
    diff_median = median_exp2 - median_exp1,
    abs_diff_median = abs(diff_median)
  ) %>%
  arrange(desc(abs_diff_median))

```


```{r plot the anchor meanings}
item_comp %>%
  ggplot(aes(x = mean_exp1, y = mean_exp2, label = as.character(item_id))) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  geom_point(size = 2) +
  geom_text(check_overlap = TRUE, nudge_y = 0.05, size = 3) +
    coord_fixed(ratio = 0.6) +
  labs(x = "Exp1 anchor mean rating", y = "Exp2 anchor mean rating",
       title = "Anchor items: Exp1 vs Exp2 (means)") +
  theme_minimal(base_size = 12)
```


```{r anchor reproducibility by list}
# Exp1 norms per item (use median as robust reference)
exp1_norms <- df_anchors %>%
  filter(anchor == "exp1") %>%
  group_by(item_id) %>%
  summarise(
    exp1_mean = mean(Response, na.rm = TRUE),
    exp1_median = median(Response, na.rm = TRUE),
    .groups = "drop"
  )

# Exp2 item means, separated by list
exp2_item_by_list <- df_anchors %>%
  filter(anchor == "exp2", !is.na(list)) %>%
  group_by(list, item_id) %>%
  summarise(
    exp2_mean = mean(Response, na.rm = TRUE),
    exp2_median = median(Response, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) %>%
  left_join(exp1_norms, by = "item_id") %>%
  mutate(
    diff_mean = exp2_mean - exp1_mean,
    diff_median = exp2_median - exp1_median
  )

```


```{r Participant-level anchor screening as no one excluded}
# Compute participant anchor deviation score in Exp2
exp2_dev <- df_anchors %>%
  filter(anchor == "exp2", !is.na(list)) %>%
  left_join(exp1_norms, by = "item_id") %>%
  mutate(abs_dev = abs(Response - exp1_median)) %>%
  group_by(subject, list) %>%
  summarise(
    n_anchors = n(),
    mean_abs_dev = mean(abs_dev, na.rm = TRUE),
    max_abs_dev = max(abs_dev, na.rm = TRUE),
    n_dev_ge3 = sum(abs_dev >= 3, na.rm = TRUE),
    .groups = "drop"
  )


# exclude if ≥3 of the 5 anchors were rated 1–3
exp2_low_anchor_rule <- df_anchors %>%
  filter(anchor == "exp2", !is.na(list)) %>%
  group_by(subject, list) %>%
  summarise(
    n_anchors = n(),
    n_low = sum(Response %in% c(1,2,3,4), na.rm = TRUE),
    fail_low_rule = (n_low >= 3),
    .groups = "drop"
  )

```


```{r report}
# the output of the per-list summary
exp2_item_by_list %>%
  group_by(list) %>%
  summarise(
    mean_abs_diff_median = mean(abs(diff_median), na.rm = TRUE),
    mean_abs_diff_mean = mean(abs(diff_mean), na.rm = TRUE),
    cor_means = cor(exp1_mean, exp2_mean, use = "complete.obs"),
    .groups = "drop"
  )

# the QC counts: Participant-level anchor deviation score (relative to Exp1 medians)
exp2_low_anchor_rule %>% count(fail_low_rule)
summary(exp2_dev$mean_abs_dev) 
```


## Experiment 2: Response distributions and reliability
### Data screening: remove extremely fast responses (RT < 500 ms)
We first remove subject 12a who responded extremely fast on many trials (10.7% of total RTs).
The histogram of 12a's Response is also weird.
```{r remove the rows whose RT is less than 500ms, message=FALSE, warning=FALSE}
# set working directory to project folder (update as needed)
setwd("~/Documents/meaning-specific-AoA-of-562-Chinese-homonyms/data")

# load data
df_raw <- read_csv("Exp2rating_homonym_data.csv")

# report number and percent of rows with RT < 500 per subject, then remove them
removed_rows_report <- df_raw %>%
  group_by(file) %>%
  summarize(
    total_rows = n(),
    removed_rows = sum(RT_ms < 500, na.rm = TRUE),
    removed_pct = (removed_rows / total_rows) * 100,
    .groups = "drop"
  ) %>%
  filter(removed_rows > 0) %>%
  arrange(desc(removed_rows))

print(removed_rows_report)

# define file is subj
subj <- "cleaned_12a"

gghistogram(df_raw %>% filter(file== subj), x="Response", bins=7)
ggscatter(df_raw %>% filter(file== subj), x="item_id", y="RT_ms") + ylim(0, 10000)


# remove 12a
df_raw <- df_raw %>% filter(file != "cleaned_12a")

```

### norms and response distributions
```{r norms}
# preprocess the raw data
df_raw <- df_raw %>%
  mutate(
    file = as.character(file),
    subject = as.character(subject),
    item_id = as.integer(item_id),
    Response = suppressWarnings(as.integer(Response)),
    is_anchor = item_id == 9998
  ) %>%
  # Parse list number and part letter from strings like "cleaned_10a"
  mutate(
    list_num = suppressWarnings(as.integer(str_match(file, "cleaned_(\\d+)([a-z])")[, 2])),
    part = str_match(file, "cleaned_(\\d+)([a-z])")[, 3],
    list = paste0("List", list_num),
    file_id = paste0(list, part)  # e.g., "List10a"
  ) %>%
  # Robust participant ID: unique within each file
  mutate(
    subject_uid = paste(file, subject, sep = "::")
  )

# Meaning-level summary table (database norms)
meaning_norms <- df_raw %>%
  filter(!is_anchor) %>%
  group_by(item_id, headword, meaning) %>%
  summarise(
    n = n(),
    mean = mean(Response),
    sd = sd(Response),
    median = median(Response),
    q1 = quantile(Response, 0.25, names = FALSE, type = 2),
    q3 = quantile(Response, 0.75, names = FALSE, type = 2),
    iqr = q3 - q1,
    se = sd / sqrt(n),
    ci95_low = mean - 1.96 * se,
    ci95_high = mean + 1.96 * se,
    .groups = "drop"
  )

# Distribution of meaning-level means and SDs (overall)
mean_summary_overall <- meaning_norms %>%
  summarise(
    mean_min = min(mean),
    mean_p10 = quantile(mean, 0.10, names = FALSE, type = 2),
    mean_median = median(mean),
    mean_p90 = quantile(mean, 0.90, names = FALSE, type = 2),
    mean_max = max(mean)
  )
mean_summary_overall

```

### plot the distribution
```{r plot RESP distribution}
# plot the RESP distribution 
df_raw %>%
  filter(!is_anchor, Response %in% 1:7) %>%
  ggplot(aes(x = factor(Response, levels = 1:7))) +
  geom_bar(width = 0.70, fill = "lightblue", color = "black") +
  scale_x_discrete(drop = FALSE, expand = expansion(add = 0.60)) +
  labs(title = "Distribution of Response values", x = "Response", y = "Frequency") +
  theme_minimal()

```


### split-half reliability
```{r Split-half reliability of meaning means}
# Split-half reliability of meaning-level means, pooling ALL files
# (anchors removed; valid responses only; subject IDs kept unique by file)

set.seed(1)

spearman_brown <- function(r) (2 * r) / (1 + r)

# --- prepare pooled, valid target trials ---
df_pool <- df_raw %>%
  mutate(
    Response = suppressWarnings(as.integer(Response)),
    item_id  = as.integer(item_id),
    is_anchor = item_id == 9998,
    resp_valid = !is.na(Response) & Response %in% 1:7,
    subject_uid = paste(as.character(file), as.character(subject), sep = "::")
  ) %>%
  filter(!is_anchor, resp_valid) %>%
  select(subject_uid, item_id, Response)

# --- one split-half draw ---
split_half_once_pooled <- function(dat) {
  subs <- unique(dat$subject_uid)
  n_subs <- length(subs)
  half_n <- floor(n_subs / 2)

  g1 <- sample(subs, half_n)
  g2 <- setdiff(subs, g1)

  m1 <- dat %>%
    filter(subject_uid %in% g1) %>%
    group_by(item_id) %>%
    summarise(m1 = mean(Response), .groups = "drop")

  m2 <- dat %>%
    filter(subject_uid %in% g2) %>%
    group_by(item_id) %>%
    summarise(m2 = mean(Response), .groups = "drop")

  joined <- inner_join(m1, m2, by = "item_id")

  # Pearson correlation of item means across halves
  r <- suppressWarnings(cor(joined$m1, joined$m2, use = "complete.obs"))

  tibble(
    n_subs = n_subs,
    n_items = nrow(joined),
    r = r,
    r_sb = spearman_brown(r)
  )
}

# --- run many random splits ---
n_iter <- 2000
split_results_pooled <- map_dfr(seq_len(n_iter), ~ split_half_once_pooled(df_pool))

# --- summarise reliability across random splits ---
split_summary_pooled <- split_results_pooled %>%
  summarise(
    iter = n(),
    n_subs = median(n_subs),
    n_items = median(n_items),
    r_median = median(r, na.rm = TRUE),
    r_sb_median = median(r_sb, na.rm = TRUE),
    r_sb_ci_low = quantile(r_sb, 0.025, na.rm = TRUE, names = FALSE, type = 2),
    r_sb_ci_high = quantile(r_sb, 0.975, na.rm = TRUE, names = FALSE, type = 2)
  )

split_summary_pooled

```


## Experiment 2: Rating latency and AoA

```{r RT analysis}
df_rt2 <- df_raw %>%
  filter(item_id != 9998) %>%                           # exclude anchors
  filter(!is.na(Response), Response %in% 1:7) %>%       # valid ratings
  filter(!is.na(RT_ms), RT_ms > 0) %>%                  # valid RTs
  mutate(
    logRT = log(RT_ms),
    trial_z = as.numeric(scale(stimuli_order))
  ) %>%
  group_by(subject) %>%
  mutate(
    logRT_med = median(logRT, na.rm = TRUE),
    logRT_mad = mad(logRT, na.rm = TRUE),               
    keep_logRT = abs(logRT - logRT_med) <= 2.5 * logRT_mad
  ) %>%
  ungroup() %>%
  filter(keep_logRT)


#Create within- and between-participant AoA-rating predictors
df_rt2 <- df_rt2 %>%
  group_by(subject) %>%
  mutate(
    RESP_mean = mean(Response, na.rm = TRUE),
    RESP_w = Response - RESP_mean
  ) %>%
  ungroup() %>%
  mutate(
    RESP_b = as.numeric(scale(RESP_mean, center = TRUE, scale = FALSE))
  )

# Mixed model (Exp2): does higher AoA rating predict longer latency?
library(lme4)
library(lmerTest)

m_exp2 <- lmer(
  logRT ~ RESP_w + trial_z + RESP_b +
    (1 + RESP_w | subject) +
    (1 | item_id),
  data = df_rt2,
  REML = TRUE
)

summary(m_exp2)
confint(m_exp2, method = "Wald")

```


```{r plot}
# Conditional (marginal) effect of RESP_w on logRT
m <- m_exp2  # the fitted model

# Build a grid for RESP_w; hold other predictors at 0 (their centered values)
x_rng <- range(df_rt2$RESP_w, na.rm = TRUE)
grid <- data.frame(
  RESP_w  = seq(x_rng[1], x_rng[2], length.out = 200),
  trial_z = 0,
  RESP_b  = 0
)

# Fixed-effects prediction and its SE via model matrix + vcov
X  <- model.matrix(~ RESP_w + trial_z + RESP_b, data = grid)
b  <- fixef(m)
V  <- as.matrix(vcov(m))

grid$fit <- as.numeric(X %*% b)
grid$se  <- sqrt(diag(X %*% V %*% t(X)))
grid$lwr <- grid$fit - 1.96 * grid$se
grid$upr <- grid$fit + 1.96 * grid$se

# add a binned mean of the observed data for intuition (still on logRT)
p <- ggplot() +
  geom_ribbon(data = grid, aes(x = RESP_w, ymin = lwr, ymax = upr),
              alpha = 0.18) +
  geom_line(data = grid, aes(x = RESP_w, y = fit), linewidth = 1) +
  stat_summary_bin(
    data = df_rt2,
    aes(x = RESP_w, y = logRT),
    bins = 30,
    fun = mean,
    geom = "point",
    size = 2,
    alpha = 0.75
  ) +
  labs(
    x = expression(paste("Within-participant AoA deviation (RESP"[w], ")")),
    y = "Predicted logRT (fixed effects) / observed binned mean",
    title = "Experiment 2: Within-participant AoA ratings do not predict rating latency"
  ) +
  theme_classic()

p

```


## Validation analyses
### Convergent validity with published word-level AoA norms
```{r test on word-level AoA norms}
# set working directory to project folder (update as needed)
setwd("~/Documents/meaning-specific-AoA-of-562-Chinese-homonyms/data")

# read the existing AoA norms from excel file
aoa_norms <- read_excel('Word AoA_Xu, X., et al. (2021).xlsx')
# rename the column "AoA Mean" to "AoA_word"
aoa_norms <- aoa_norms %>% rename(AoA_word = `AoA Mean`)
# select columns we need
aoa_norms <- aoa_norms %>% select(Word, AoA_word)
# string hygiene: trim whitespace from Word column
aoa_norms <- aoa_norms %>%
  transmute(Word = str_squish(Word),
            AoA_word)

# read our rating data
df_rating <- read_csv('Chinese_homonym_meaning_level_norms.csv', show_col_types = FALSE)
# rename the column "headword" to "Word"
df_rating <- df_rating %>% rename(Word = headword)
# calculate AoA summary statistics from the rating column named "mean"
df_rating <- df_rating %>%
  mutate(Word = str_squish(Word)) %>%
  group_by(Word) %>%
  summarise(
    AoA_mean = mean(mean, na.rm = TRUE),
    AoA_min  = min(mean,  na.rm = TRUE),
    AoA_max  = max(mean,  na.rm = TRUE),
    .groups = "drop"
  )

# merge the two dataframes by Word and remove the rows with missing values in either AoA_mean or AoA_word
df_norms_combined <- inner_join(df_rating, aoa_norms, by = "Word") %>%
  filter(!is.na(AoA_mean) & !is.na(AoA_word))
cat(glue::glue("Number of words with both AoA_mean and AoA_word: {nrow(df_norms_combined)}\n Not all homonyms in our ratings have corresponding AoA norms. The correlation will be based on this subset of words.\n"))

# check the correlation between AoA_mean and the existing AoA norms
correlation_mean1 <- df_norms_combined %>%
  select(AoA_mean, AoA_word) %>%
  na.omit() %>%
  cor.test(~ AoA_mean + AoA_word, data = .)
# print the correlation result with a message
cat("\nCorrelation between AoA_mean and AoA_word:\n")
print(correlation_mean1)

# check the correlation between AoA_min and the existing AoA norms
correlation_min <- df_norms_combined %>%
  select(AoA_min, AoA_word) %>%
  na.omit() %>%
  cor.test(~ AoA_min + AoA_word, data = .)
# print the correlation result with a message
cat("\nCorrelation between AoA_min and AoA_word:\n")
print(correlation_min)

# check the correlation between AoA_max and the existing AoA norms
correlation_max <- df_norms_combined %>%
  select(AoA_max, AoA_word) %>%
  na.omit() %>%
  cor.test(~ AoA_max + AoA_word, data = .)
# print the correlation result with a message
cat("\nCorrelation between AoA_max and AoA_word:\n")
print(correlation_max)

```


### Predictive validity: meaning-level AoA beyond word-level AoA
```{r data structure, message=FALSE}
# set working directory to project folder (update as needed)
setwd("~/Documents/meaning-specific-AoA-of-562-Chinese-homonyms/data")

# load the data from a CSV file
df_judgment_aoa <- read_csv("Homonym disambiguation judgments_(Xie et al., 2025).csv")

# Coding and preprocessing
df <- df_judgment_aoa %>%
  mutate(
    Subject = factor(Subject),
    Word = factor(Word),
    SentencePair_id = factor(SentencePair_id),
    Sense   = factor(Sense,   levels = c("Same","Different")),
    Context = factor(Context, levels = c("Same","Different")),
    POSWord = factor(POSWord, levels = c("Same","Different"))
  )

# sum-to-zero contrasts (recommended for factorial designs)
contrasts(df$Sense)   <- contr.sum(2)  # Same = 1; Different = -1
contrasts(df$Context) <- contr.sum(2)  # Same = 1; Different = -1
contrasts(df$POSWord) <- contr.sum(2)  # Same = 1; Different = -1

# RT analysis dataset: correct trials only
df_rt <- df %>%
  filter(correctness == 1) %>%
  mutate(
    logRT = log(Stimulus_RT),

    Trial_z = as.numeric(scale(Trial)),
    Word_logW_z = as.numeric(scale(Word_logW)),

    AoA_word_z = as.numeric(scale(AoA_word)),
    AoA_mean_z = as.numeric(scale(AoA_rating_mean)),
    AoA_diff_z = as.numeric(scale(AoA_rating_diff_abs))
  )
```
We used sum coding for all binary experimental factors (Sense, Context, POSWord) such that main effects reflect averages across the other factors.

```{r modeling}
# Primary RT models: crossed random intercepts for Subject and SentencePair
m0 <- lmer(
  logRT ~ Sense*Context*POSWord + Trial_z + Word_logW_z +
    (1 | Subject) + (1 | SentencePair_id),   # (1 | Subject) accounts for participant speed
                                             # (1 | SentencePair_id) accounts for sentence-pair difficulty
  data = df_rt, REML = FALSE
)

m_word <- update(m0, . ~ . + AoA_word_z)

# explicitly test nested steps one-by-one
m_mean_only <- update(m0, . ~ . + AoA_mean_z)
m_diff_only <- update(m0, . ~ . + AoA_diff_z)

anova(m0, m_mean_only)
anova(m0, m_diff_only)

m_word_plus_mean <- update(m_word, . ~ . + AoA_mean_z)
m_word_plus_diff <- update(m_word, . ~ . + AoA_diff_z)
m_word_plus_both <- update(m_word, . ~ . + AoA_mean_z + AoA_diff_z)

anova(m_word, m_word_plus_mean)
anova(m_word, m_word_plus_diff)
anova(m_word, m_word_plus_both)

```
AoA_mean_z is clearly > 0 (slower RT for later-acquired meanings), AoA_diff_z is not. 
Adding AoA_word_z does not help, and it is not significant in the full model, 
so we will keep the simpler meaning-level AoA model without the word-level AoA predictor (good for interpretability and parsimony).

Including (1|SentencePair_id) controls for many nuisance properties that otherwise inflate AoA effects:
(1)sentence length/complexity
(2)plausibility/naturalness
(3)specific lexical neighbors in the sentences
(4)local cues that make the sense easy/hard
This makes any AoA effect much easier to defend as meaning/lexicon-related, not sentence-artifact-related.

```{r test interactions}
# Targeted interactions (optional, but theoretically motivated)
m_word_plus_mean_int <- update(
  m_word_plus_mean,
  . ~ . + AoA_mean_z:Sense
)

anova(m_word_plus_mean, m_word_plus_mean_int)
```
No support for AoA × Sense moderation, so keep the simpler additive AoA model (good).

```{r robustness checks, message=FALSE}
# Add Word as an additional random intercept for testing robustness to word-level clustering (beyond sentence-pair clustering).
# Statistically, Word is a higher-level grouping above SentencePair_id (pairs are nested in word). 
# Adding (1|Word) can capture any word-form–specific variance shared across the 8 pairs for that word beyond what the sentence-pair intercepts capture.
# However, because each pair already has its own intercept, (1|Word) often adds little and can be redundant (or lead to near-zero variance estimates).
m_both_wordRE <- lmer(
  logRT ~ Sense*Context*POSWord + Trial_z + Word_logW_z +
    AoA_word_z + AoA_mean_z + AoA_diff_z +
    (1 | Subject) + (1 | SentencePair_id) + (1 | Word),
  data = df_rt, REML = FALSE
)

summary(m_both_wordRE)


# By-subject random slope for Sense (if it converges cleanly)
m_both_rs <- lmer(
  logRT ~ Sense*Context*POSWord + Trial_z + Word_logW_z +
    AoA_word_z + AoA_mean_z + AoA_diff_z +
    (1 + Sense | Subject) + (1 | SentencePair_id),
  data = df_rt, REML = FALSE,
  control = lmerControl(optimizer = "bobyqa")
)

summary(m_both_rs)
```
The Word intercept variance is non-zero (Var ≈ 0.0059), so there is measurable word-level clustering beyond sentence-pair and subject. 
Therefore, treat (1|Word) as a robustness model, not necessarily the primary one. 
The random-slope model (1 + Sense | Subject) fits much better than the random-intercept-only model, 
so we will re-run the same nested AoA tests there and report those as primary.


```{r final model}
# final fixed-effects AoA specification
m_final <- lmer(
  logRT ~ Sense*Context*POSWord + Trial_z + Word_logW_z +
    AoA_word_z + AoA_mean_z +
    (1 | Subject) + (1 | SentencePair_id),
  data = df_rt, REML = FALSE
)

# document that adding AoA_diff_z does not help
m_final_plus_diff <- update(m_final, . ~ . + AoA_diff_z)
anova(m_final, m_final_plus_diff)

# Because random-slope model (1 + Sense | Subject) fit much better earlier, 
# re-run the same nested AoA tests there and report those as primary
m_rs0 <- lmer(
  logRT ~ Sense*Context*POSWord + Trial_z + Word_logW_z +
    (1 + Sense | Subject) + (1 | SentencePair_id),
  data = df_rt, REML = FALSE,
  control = lmerControl(optimizer="bobyqa")
)

m_rs_word <- update(m_rs0, . ~ . + AoA_word_z)
m_rs_word_mean <- update(m_rs_word, . ~ . + AoA_mean_z) # Recommended “final” model to take forward
m_rs_word_mean_diff <- update(m_rs_word_mean, . ~ . + AoA_diff_z)

anova(m_rs0, m_rs_word, m_rs_word_mean, m_rs_word_mean_diff)
summary(m_rs_word_mean)

# Convert slope to % RT change per 1-point rating
100 * (exp(fixef(m_rs_word_mean)[c("AoA_word_z", "AoA_mean_z")]) - 1)

```
Under the better random-effects structure (1 + Sense | Subject), the story is the same—but stronger and cleaner
Interpretation: both word-level AoA and meaning-level AoA (mean) explain RT variation; AoA difference between meanings does not.


